Viscous Erosion in Porous Media

### 2018-02-02 docs/methods/erosion_methods.tex
small changes to method paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, several flows can be described
to good accuracy under a two-dimensional approximation, certain
phenomena occur only in three dimensions.  Extending the current work to
three dimensions is possible with a boundary integral equation
formulation.  In particular, the fluid solvers exist~\cite{mal-bir2015},
and the main challenges of a BIE solver have been addressed by other
groups including surface representation~\cite{yin-bir-zor2006,
vee-rah-bir-zor2011}, near-singular integration~\cite{yin-bir-zor2006,
kli-tor2016b}, and fast summation methods~\cite{yin-bir-zor2004,
ros-ols2015}.  The more significant challenge of extending our work to
three dimensions is guaranteeing that the mesh does not become
distorted, which can be done, for example, by using an extension of the
two-dimensional {\thL} formulation~\cite{amb-sie-tlu2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We only
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies.  The bodies are denoted as $\gamma_\ell$,
$\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup
\gamma_M$.  Neglecting inertial forces, the dynamics of the fluid is
fully characterized by the position of the bodies $\xx_\ell(s,t) \in
\gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribed a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.  The outer wall is formed by rounding off the corners
of $[-3,3] \times [-1,1]$, and we enforce the Hagen-Poiseuille flow
\begin{align*}
  \UU(\xx) = \umax (1-y^2,0)
\end{align*}
on $\Gamma$.  Here, $U$ sets the maximum of the imposed Poiseuille flow. Often, we will simply set $U=1$, but in some instances we will change $U$ dynamically to enforce that the pressure drop across the computational cell is constant as bodies erode within it.
The boundary of the fluid domain is $\bd\Omega = \gamma_1
\cup \cdots \cup \gamma_M \cup \Gamma$.  To minimize the boundary
effects of the inflow and outflow, we only place bodies in the center
third, $[-1,1] \times [-1,1]$, of the fluid domain.
\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing
    equations.  A no-slip boundary condition is imposed on each body
    $\gamma_\ell$ whose unit normal points outward relative to the
    geometry.  On the outer geometry $\Gamma$, a pipe flow is imposed.
    The bodies are eroded at a rate that is proportional to the
    magnitude of the shear stress applied by the Stokesian fluid.  The
    bodies are constrained to the middle third of the channel that is
    located between the dashed lines.}
\end{figure}

We use a quasi-static approximation.  That is, at each time step, we
freeze the geometry, solve the incompressible Stokes equations, and then
update the geometry according to the erosion model.  A flow diagram is
illustrated in Figure~\ref{fig:workflow}.  The two steps inside the red
rectangle describes a time step in the quasi-static approximation
\begin{figure}[htpb]
  \centering
  \input{figs/workflow.tikz}
  \caption{\label{fig:workflow}A caption.}
\end{figure}

@@

To model erosion, the normal velocity is prescribed to be
proportional to the absolute value of the shear stress.   Since the
constant of proportionality only sets the time scale, we let $\Vn =
|\tau|$ as described in equation~\eqref{eqn:erosionModel}.
\begin{align}
  \Vn = \abs{\tau} \nn.
  \label{eqn:bodyEvolution}
\end{align}
For the sake of numerical stability, we modify this law in
Section~\ref{sec:thetaL}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
\label{sec:bies}
Here we describe a boundary integral equation to solve the fluid
equations. In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
As the viscosity $\mu$ also sets the time scale, we take $\mu=1$.  We
start by defining the double-layer potential
$\eeta:\bd\Omega \rightarrow \RR^2$,
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, $\nn$ is the unit outward
normal at $\yy$, and $\eeta$ is an unknown density function.  The
double-layer potential can not capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}, and this is rectified by
introducing additional terms for each body.  The rank deficiency removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint that results from the flux-free requirement of $\ff$.

@@

For the sake of numerical stability, we modify the erosion law in
equation~\eqref{eqn:bodyEvolution} to include a term that depends on the
local curvature $\kappa$. The idea is to prevent regions of very high
curvature. The modified normal velocity is 
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{align}
Here, $\eps$ is a small parameter, $L$ is the total arc length of the
body, and $\mean{\cdot}$ indicates a boundary mean. The second term on
the right acts to smooth bodies with high-frequency oscillations by
introducing a {\em curvature}-dependent component in the interface
velocity. Since a closed body has a mean curvature of $\mean{\kappa} =
2\pi / L$, the term in parentheses has mean zero. Thus, this term is
area-preserving, so that the only material loss is due to the shear
stress itself. In addition, we have scaled this term on the
$\mean{\atau}$, the average absolute stress, so that it is always
proportional to the true source of erosion $\atau$, even as bodies
vanish. \todo[inline]{reword}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, several flows can be described
to good accuracy under a two-dimensional approximation, but certain
phenomena occur only in three dimensions.  Extending the current work to
three dimensions is possible with a boundary integral equation
formulation.  In particular, the fluid solvers exist~\cite{mal-bir2015},
and the main challenges of a BIE solver have been addressed by other
groups including surface representation~\cite{yin-bir-zor2006,
vee-rah-bir-zor2011}, near-singular integration~\cite{yin-bir-zor2006,
kli-tor2016b}, and fast summation methods~\cite{yin-bir-zor2004,
ros-ols2015}.  The more significant challenge of extending our work to
three dimensions is guaranteeing that the mesh does not become
distorted, which can be done, for example, by using an extension of the
two-dimensional {\thL} formulation~\cite{amb-sie-tlu2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We only
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies.  The bodies are denoted as $\gamma_\ell$,
$\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup
\gamma_M$.  Neglecting inertial forces, the dynamics of the fluid is
fully characterized by the position of the bodies $\xx_\ell(s,t) \in
\gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribe a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{\em conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{\em no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{\em outer wall
velocity} \\
\Vn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{\em erosion model},
\end{split}
\end{equation}
where $\Vn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.  The outer wall is formed by rounding off the corners
of $[-3,3] \times [-1,1]$, and we enforce the Hagen-Poiseuille flow
\begin{align*}
  \UU(\xx) = \umax (1-y^2,0), \quad \xx \in \Gamma,
\end{align*}
where $U$ sets the maximum of the imposed Poiseuille flow. Often, we
will simply set $U=1$, but in some instances we change $U$ dynamically
to enforce that the pressure drop across the computational cell is
constant as bodies erode within it.  The boundary of the fluid domain is
$\bd\Omega = \gamma_1 \cup \cdots \cup \gamma_M \cup \Gamma$.  To
minimize the boundary effects of the inflow and outflow, we only place
bodies in the center third, $[-1,1] \times [-1,1]$, of the fluid domain.
\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing
    equations.  A no-slip boundary condition is imposed on each body
    $\gamma_\ell$ whose unit normal points outward relative to the
    geometry.  On the outer geometry $\Gamma$, a Hagen-Poiseuille flow
    is imposed.  The bodies are eroded at a rate that is proportional to
    the magnitude of the shear stress applied by the Stokesian fluid.
    The bodies are constrained to the middle third of the channel that
    is located between the dashed lines.}
\end{figure}

We use a quasi-static approximation.  That is, at each time step, we
freeze the geometry, solve the incompressible Stokes equations, and then
update the geometry according to the erosion model.  The two steps
inside the red rectangle describes a time step in the quasi-static
approximation.
%A flow diagram is illustrated in Figure~\ref{fig:workflow}.  
%\begin{figure}[htpb]
%  \centering
%  \input{figs/workflow.tikz}
%  \caption{\label{fig:workflow}A caption.}
%\end{figure}

@@

To model erosion, the normal velocity is  proportional to the absolute
value of the shear stress.   Since the constant of proportionality only
sets the time scale, we let $\Vn = |\tau|$ as described in
equation~\eqref{eqn:erosionModel}. For the sake of numerical stability,
we modify this law in Section~\ref{sec:thetaL}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Boundary Integral Equation Formulation} 
\label{sec:bies}
Here we describe a boundary integral equation to solve the fluid
equations. In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the grains $\gamma$.  As the viscosity $\mu$
also sets the time scale, we take $\mu=1$.  We start by defining the
double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, $\nn$ is the unit outward
normal at $\yy$, and $\eeta$ is an unknown density function.  The
double-layer potential can not capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}.  The rank deficiency is removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint that results from the flux-free requirement of $\ff$.

@@

For the sake of numerical stability, we modify the erosion law in
equation~\eqref{eqn:erosionModel} to include a term that depends on the
local curvature $\kappa$. The idea is to prevent regions of very high
curvature. The modified normal velocity is 
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{align}
Here, $\eps$ is a small parameter, $L$ is the total arc length of the
body, and $\mean{\cdot}$ indicates a boundary mean. The second term on
the right acts to smooth bodies with high-frequency oscillations by
introducing a {\em curvature}-dependent component in the interface
velocity. Since a closed body has a mean curvature of $\mean{\kappa} =
2\pi / L$, the term in parentheses has mean zero. Thus, this term is
area-preserving, so that the only material loss is due to the shear
stress itself. In addition, we have scaled this term on the
$\mean{\atau}$, the average absolute stress, so that it is always
proportional to the true source of erosion $\atau$, even as bodies
vanish. \todo[inline]{reword}
